divan: An R Shiny app for exploratory functional diversity analyses

### 2023-11-24 paper/paper.Rmd
update ms

---
title: 'divan: An R Shiny app for exploratory functional diversity analyses'
tags:
- Shiny
- functional ecology
- biodiversity
- functional traits
authors:
- name: Facundo X. Palacio
  affiliation: 1, 2
  corresponding: yes
- name: "Caio Graco-Roza"
  affiliation: 3
- name: Emma J. Hudgins
  affiliation: 4
date: "2 November 2023"
affiliations:
- name: Sección Ornitología, Universidad Nacional de La Plata, Argentina
  index: 1
- name: Consejo Nacional de Investigaciones Científicas y Técnicas, Argentina
  index: 2
- name: University of Helsinki, Helsinki, Finland
  index: 3
- name: School of Agriculture, Food, and Ecosystem Sciences, The University of Melbourne,
    Australia
  index: 4
bibliography: paper.bib
output: md_document
---

@@

# Overview
The app layout is split into a sidebar, through which the user can navigate across tabs with the different steps of an exploratory functional diversity analysis, and a main window within which the user can visually explore and interact with the different outputs (\autoref{fig:fig1}). Once the app is running, users can begin the workflow by loading two files: (1) a site by species dataframe (“community matrix”), the usual type of dataset used in classical community ecology, and (2) a species by trait dataframe (“trait matrix”) containing the information about species traits. After loading these two files, numerical and graphical summaries for each dataset are provided in the next submenu down along the sidebar. Beyond this, further sidebar submenus allow for the creation of several plots. The community matrix can be visually explored through a species by site heatmap with dendrogram, individual-based rarefaction curves, rank-abundance curves and species richness and prevalence (proportion of sampling units occupied by each species) histograms. The trait matrix can be visually explored through univariate plots, including boxplots, histograms and kernel density functions. The app also allows the user to assess collinearity between traits through Pearson’s correlations, visual inspection of scatterplots and linear or non-linear model fits, and automatically identifying missing data. Finally, the user can save all the plots produced as a single PDF file for later visualization.

![The graphical user interface (GUI) of **divan** after loading a community matrix. (a) Sidebar menu, (b) heatmap plus dendrogram of the community matrix, (c) rarefaction curves of individual sites. \label{fig:fig1}](../fig/fig_1.jpg)

After exploring summaries of the community and trait data, the next step involves constructing a functional trait space (subspace occupied by a set of species in a lower dimensional Euclidean space, in which each axis or dimension represents a functional trait). One of the main benefits of the app is that this space can be built following different approaches. Functional dendrograms use hierarchical clustering [@Petchey2002; @BAT], and allows the user delimiting clusters of species playing similar ecological roles (functional groups; @Walker1992) based on the numerous available algorithms in the package NbClust [@NbClust]. Functional ordination is based on Principal Coordinate Analysis (PCoA), and allows the user selecting different dissimilarity metrics (Euclidean, Gower, Manhattan, Mahalanobis). The reduced space is visualized in two dimensions, along with a convex hull encompassing all the species of the sampling unit and a 2D kernel density estimate (\autoref{fig:fig2}). Information about the PCoA, including percentage of explained variability and a screeplot of the different dimensions, is also provided. The last approach to build functional trait spaces is based on probabilistic hypervolumes [@Blonder2018a; @Blonder2018b], which represent a density-based estimation of the multidimensional trait space, based on the distribution of the species (or any other taxonomic entity) observed therein [@Mammola2020]. The resulting output is a plot of the first two dimensions depicting the species, the random points and the predicted boundaries of the hypervolumes for each sampling unit.

![The GUI of divan after loading a trait matrix. (a) Sidebar menu, (b) selection of functional traits for analysis and plotting, (c) Principal Coordinate Analysis (PCoA) inputs, (d) biplot of species and functional traits plus density kernel surface.\label{fig:fig2}](../fig/fig_2.jpg)

@@

Requirements and limitations
In principle, the package needs both data matrices (i.e., community and trait matrices) to fully operate (i.e,. computing functional diversity). However, both matrices are independent from each other, so that the user can explore the community matrix on the one hand (e.g., if the researcher is interested in describing species assemblages), and the trait data table on the other hand (e.g., if the study aims to search for patterns in trait spaces).

>>>>

---
title: 'divan: An R Shiny app for exploratory functional diversity analyses'
tags:
- Shiny
- functional ecology
- biodiversity
- functional traits
authors:
- name: Facundo X. Palacio
  affiliation: 1, 2
  corresponding: yes
  email: facundo_palacio@fcnym.unlp.edu.ar
- name: "Caio Graco-Roza"
  affiliation: 3
- name: Emma J. Hudgins
  affiliation: 4
date: "17 November 2023"
affiliations:
- name: Sección Ornitología, Universidad Nacional de La Plata, Argentina
  index: 1
- name: Consejo Nacional de Investigaciones Científicas y Técnicas, Argentina
  index: 2
- name: University of Helsinki, Helsinki, Finland
  index: 3
- name: School of Agriculture, Food, and Ecosystem Sciences, The University of Melbourne,
    Australia
  index: 4
bibliography: paper.bib
output: md_document
---

@@

# Overview
The app layout is split into a sidebar, through which the user can navigate across tabs with the different steps of an exploratory functional diversity analysis, and a main window within which the user can visually explore and interact with the different outputs (\autoref{fig:example}). Once the app is running, users can begin the workflow by loading two files: (1) a site by species dataframe (“community matrix”), the usual type of dataset used in classical community ecology, and (2) a species by trait dataframe (“trait matrix”) containing the information about species traits. After loading these two files, numerical and graphical summaries for each dataset are provided in the next submenu down along the sidebar. Beyond this, further sidebar submenus allow for the creation of several plots. The community matrix can be visually explored through a species by site heatmap with dendrogram, individual-based rarefaction curves, rank-abundance curves and species richness and prevalence (proportion of sampling units occupied by each species) histograms. The trait matrix can be visually explored through univariate plots, including boxplots, histograms and kernel density functions. The app also allows the user to assess collinearity between traits through Pearson’s correlations, visual inspection of scatterplots and linear or non-linear model fits, and automatically identifying missing data. Finally, the user can save all the plots produced as a single PDF file for later visualization.

![The graphical user interface (GUI) of **divan** after loading a community matrix. (a) Sidebar menu, (b) heatmap plus dendrogram of the community matrix, (c) rarefaction curves of individual sites.\label{fig:example}](../fig/fig_1.jpg)

After exploring summaries of the community and trait data, the next step involves constructing a functional trait space (subspace occupied by a set of species in a lower dimensional Euclidean space, in which each axis or dimension represents a functional trait). One of the main benefits of the app is that this space can be built following different approaches. Functional dendrograms use hierarchical clustering [@Petchey2002; @BAT], and allows the user delimiting clusters of species playing similar ecological roles (functional groups; @Walker1992) based on the numerous available algorithms in the package NbClust [@NbClust]. Functional ordination is based on Principal Coordinate Analysis (PCoA), and allows the user selecting different dissimilarity metrics (Euclidean, Gower, Manhattan, Mahalanobis). The reduced space is visualized in two dimensions, along with a convex hull encompassing all the species of the sampling unit and a 2D kernel density estimate (\autoref{fig:example_2}). Information about the PCoA, including percentage of explained variability and a screeplot of the different dimensions, is also provided. The last approach to build functional trait spaces is based on probabilistic hypervolumes [@Blonder2018a; @Blonder2018b], which represent a density-based estimation of the multidimensional trait space, based on the distribution of the species (or any other taxonomic entity) observed therein [@Mammola2020]. The resulting output is a plot of the first two dimensions depicting the species, the random points and the predicted boundaries of the hypervolumes for each sampling unit.

![The GUI of divan after loading a trait matrix. (a) Sidebar menu, (b) selection of functional traits for analysis and plotting, (c) Principal Coordinate Analysis (PCoA) inputs, (d) biplot of species and functional traits plus density kernel surface.\label{fig:example_2}](../fig/fig_2.jpg)

@@

# Requirements and limitations
In principle, the package needs both data matrices (i.e., community and trait matrices) to fully operate (i.e,. computing functional diversity). However, both matrices are independent from each other, so that the user can explore the community matrix on the one hand (e.g., if the researcher is interested in describing species assemblages), and the trait data table on the other hand (e.g., if the study aims to search for patterns in trait spaces).
